---
title: "sge2_behavior_methods_script.Rmd"
output: html_document
date: "2022-10-20"
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(data.table)
library(viridis)
library(cowplot)
library(grid)
library(EloRating)

workDir <- "~/Dropbox (Personal)/macaque/vaccine/"
dataDir <- paste0(workDir,"data/")
modelFolder <- paste0(workDir,"models/")
outDir <- paste0(workDir,"output/")
figDir <- paste0(workDir,"figures/")

sampleDates <- read.table("~/Dropbox (Personal)/macaque/sge_behavior/samplingDatesForR.csv", sep = ",", header = T)
sampleDatesP2 <- read.table(paste0("~/Dropbox (Personal)/macaque/sge_behavior/data/eventDatesPhase2.csv"), sep = ",", header = T)
sampleDatesP2$group <- gsub(pattern = "NO", replacement = "", sampleDatesP2$group)
```

# Read Me
This is final code to generate group-centered Agonisms Received Rate, Overall Grooming Rate, and Elo for SGE2 phases 1 & 2.
This code will output behavioral values for all animals, for all 4 experiments (SGE2p1 Environmental Challenge, SGE2p1 Single Cell, SGE2p1 Vaccine, SGE2p2 Vaccine)
It will also generate dataset specific numbers to include with a clear methods section (behavior_methods_text.txt)


```{r cars}

```